%%% Stephen Moye
%%% Stephen_Moye@brown.edu
%%% Brown University
%%% Graphic Services

\newdimen\basesize
\newdimen\dist
\newdimen\baseblsk
\newdimen\xheight
\newdimen\ascheight
\newdimen\leading
\newdimen\adjust
\newcount\scratch
\newcount\scppt \scppt=65536
\newcount\intgp
\newcount\sscratch
\newcount\fracp

\hsize7.5in \hoffset-.5in
\vsize10in \voffset-.5in

\nopagenumbers
\parindent0pt

\def\setuptype#1#2#3#4{\font\bigtext=#3 at 18pt
\basesize=#2
\font\tinynote=#4 at 8pt
\font\bigdisp=#1 at \basesize
\font\meddisp=#1 at .75\basesize
\font\smalldisp=#1 at .5\basesize
\font\tinydisp=#1 at .41666\basesize
\setbox0=\vbox{\bigdisp ILyg}
\baseblsk=\ht0
\advance\baseblsk\dp0
\baseblsk=1.2\baseblsk}

%%% Leading calculations
\def\setleading{\setbox0=\hbox{x}
\xheight=\ht0
\setbox0=\hbox{Ilpgy}
\ascheight=\ht0
\leading=\ht0
\adjust=\xheight \advance\adjust-.5\leading
\advance\leading\dp0
\leading=1.175\leading  
\ifdim\dp0<.05\ht0 \advance\adjust.5\xheight\fi
\advance\leading by \adjust
\baselineskip\leading}

\def\prettypt#1{\dist=#1
\multiply\dist10
\scratch=\number\dist
\divide\scratch\scppt
\intgp=\scratch \divide\intgp10
\sscratch=\intgp 
\multiply\sscratch10 \advance\scratch-\sscratch
\fracp=\scratch
\hbox{\the\intgp.\the\fracp}pt\ }

%%%

\setuptype{uaur}{24pt}{Helvetica}{HelveticaI}

\leftline{\bigtext Audubon Script}

\smallskip

\hrule
\smallskip
\rightline{\tinynote Type used in James Audubon's ``Birds of America'', translated to PostScript by Richard
Beatty}
\bigskip

\line{%
\vtop to 9in{\bigdisp \baselineskip=.75\baseblsk plus 1fil
\halign{&#\hfill\cr%
Abound\cr
Bridges\cr
Canoe\cr
Dismay\cr
English\cr
French\cr
Genoa\cr
Happy\cr
Igloo\cr
Jury\cr
Kimono\cr
Lyric\cr
Musical\cr
Narcissus\cr
Opera\cr
Precise\cr
Quartz\cr
Royal\cr
Sable\cr
Transience\cr
Umbrage\cr
Velvet\cr
Waxen\cr
Xanadu\cr
Yucca\cr
Zabaglione\cr}}\hfill
\vtop to 9in{\bigdisp \baselineskip\baseblsk plus 1fil
	\halign{
 	&\hfil\thinspace#&\thinspace#\hfil&
  \enskip#\enskip&
  \hfil\thinspace#&\thinspace#\hfil\cr
  A&a&&B&b\cr
  C&c&&D&d\cr
  E&e&&F&f\cr
  G&g&&H&h\cr
  I&i&&J&j\cr
  K&k&&L&l\cr
  M&m&&N&n\cr
  O&o&&P&p\cr
  Q&q&&R&r\cr
  S&s&&T&t\cr
  U&u&&V&v\cr
  W&w&&X&x\cr
  Y&y&&Z&z\cr
   \span0&\span1\cr
   \span2&\span3\cr
   \span4&\span5&\cr
   \span6&\span7&\cr
   \span8&\&&9&\cr}}\hfill
\vtop{\hsize3in  \emergencystretch2em
\bigdisp \setleading
Brick quiz whangs jumpy veldt fox.
Nymphs vex, beg quick fjord waltz.
Quick wafting zephyrs vex bold Jim.


\rightline{\tinynote \prettypt{\basesize}/\prettypt{\the\baselineskip}}
\smallskip
\hrule

\bigskip\bigskip

\meddisp \setleading
West quickly gave Bert handsome prizes for six juicy plums.
Freight to me sixty dozen quart jars and twelve black pans.
My help squeezed back in again and joined the weavers after six.
Turgid saxophones blew over Mick's jazzy quiff.


\rightline{\tinynote \prettypt{.75\basesize}/\prettypt{\the\baselineskip}}
\smallskip
\hrule

\bigskip\bigskip

\smalldisp \setleading
Five wine experts jokingly quizzed sample chablis.
My grandfather picks up quartz and valuable onyx jewels.
Mix Zapf with Veljovic and get quirky Beziers.
All questions asked by five watch experts amazed the judge.
Back in June we delivered oxygen equipment of the same size.
We have just quoted on nine dozen boxes of grey lamp wicks.
A large fawn jumped quickly over white zinc boxes.
The exodus of jazzy pigeons is craved by squeamish walkers.

\rightline{\tinynote \prettypt{.5\basesize}/\prettypt{\the\baselineskip}}
\smallskip
\hrule

\bigskip\bigskip

\tinydisp \setleading
Murky haze enveloped a city as jarring quakes broke fourty-six windows.
The brazen jackal querulously attacked a ferral vixen, maiming her paw.
Karate experts bring down victims with just a few amazingly quick chops.
The fabled quetzal roosted on a mulberry twig, vexing the peevish jockey.
King Alexander was just partly overcome after quizzing Diogenes in his tub.
The exiled queen justly moved, celebrated with a dazzling fireworks display.

\rightline{\tinynote \prettypt{.41666666\basesize}/\prettypt{\the\baselineskip}}
\smallskip
\hrule}}

\bye



